---
title: "Comparing GCAM soil organic carbon predictions to empirical land use results"
output: bookdown::html_document2
date: "`r format(Sys.time(), '%B %d %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Methods

To begin the analysis, SOC data was taken from a basic run of GCAM, called a reference scenario. This models “business as usual” for the world, without any extra climate policies and assuming a certain level of overall warming. This GitHub repository contains all relevant data from this reference scenario used in our analysis, which was performed using R. Note that the SOC data, GCAM's soil timescale list, as well as both the geopolitical regions list and the land use regions list are separate files, and must be merged before any additional steps can be performed. Below is the code used to load the data and libraries, as well as to merge the four files listed above. The final product is a data set where each SOC data point has an associated timescale value, land use region, and geopolitical region.

```{r Read-In-GCAM-Data, message=FALSE}
library(dplyr)
library(ggplot2)
library(metafor)
library(tidyr)


#GCAM's SOC data
soilC <- read.csv(file = 'Data/GCAM_soilC.csv')

#GCAM's soil timescale list
read.csv(file = 'Data/soil_timescales.csv') %>%
  mutate(GCAM_region_ID = seq(1,32,by=1))-> timescales

#GCAM's land use region list 
glus <- read.csv(file = 'Data/GLU_codes.csv')

#GCAM's geopolitical region list
regions <- read.csv('Data/GCAM_regions.csv')

#Merging the four above data sets
soilC %>%
  mutate(GLU_code = GLU) %>%
  right_join(glus, by='GLU_code') %>%
  right_join(regions, by='GCAM_region_ID') %>%
  right_join(timescales, by='GCAM_region_ID')-> soilC_regions
```

Much of the data contained within the SOC data set is irrelevant to this analysis, so it was removed.

```{r Select-relevant-data}
soilC_regions %>%
  select(Land_Type, soil_c, GLU_code, soilTimeScale, GCAM_region_ID, Basin_long_name) -> simple_soilC_regions
```

The experimental data compared to GCAM's outputs was taken from Post & Kwon (2000) and Wei et al. (2014), both reviews of  SOC measurements from the past 70 years. This data was manually digitized into a CSV file for analysis. This data did not explicitly list site locations for each measurement, so this information was determined from the paper referenced by the two studies. Once the location was determined, it was compared to GCAM's 384 land use regions using the rmap tool and GCAM's basin code mapping file. Then, the GLU code was manually added to the digitized data from Post & Kwon.  

Once fully digitized, the experimental data was loaded into R as well. Note that there were several entries with missing timeframes or where the location of the sample could not be determined. These were assigned a value of 'NA'. 

```{r Read-In-Experimental-Data}
PostKwon <- read.csv(file= 'Data/Experimental Data.csv', na.strings = c("", "NA"))
Wei <- read.csv(file= 'Data/Wei et al Data.csv', na.strings = c("", "NA"))

```

Once the experimental data was loaded in, each observation was mapped to is corresponding GCAM data point. This was done by simply adding the previously created GCAM data set to the experimental one by land use region, geopolitical region, and land use. This process was done twice--once for the initial land use and once for the final land use. This meant that, ultimately, each experimental data point had two associated GCAM SOC values--one for the initial land use, and one for the final land use.  

This then allowed us to compute the GCAM SOC rate during land use change. This was done by simply subtracting the initial SOC value from the final value and dividing that by GCAMs soil timescale parameter. Finally, the k value (explained in depth in Wei et al.) was computed for both the experimental and GCAM data. The experimental and GCAM k values were computed using equations 1 and 2, respectively, below.
\begin{equation}
\tag{1}
k = -\frac{ln \left(\frac{C_f}{C_0}\right)}{t}
\end{equation}

\begin{equation}
\tag{2}
k = -\frac{ln (Rate \times t + 1)}{t}
\end{equation}

\newline
Experimental data from Wei et al. did not contain any raw rate information. Instead, the study only reported percent decrease in SOC concentration. This means we were only able to compare k values between the Wei et. al. data and GCAM. Additionally, k had to be computed with a modified equation, 3, below. 

\begin{equation}
\tag{3}
k = -\frac{ln \left(\frac{1}{\frac{|OC Decrease|}{100} + 1}\right)}{t}
\end{equation}

The code for merging the GCAM and experimental, as well as for computing the rate and k value is below. The process was the same for the Wei et al. data. 

```{r create_comparison_Post_Kwon}
PostKwon %>%
  select(Initial_Land_Use, Final_Land_Use, GLU_code, GCAM_region_ID, Time, Exp_Rate) %>%
  na.omit() %>%
  mutate(Land_Type = Initial_Land_Use) %>%
  right_join(simple_soilC_regions, by = c('GLU_code', 'Land_Type', 'GCAM_region_ID')) %>%
  rename(initial_soil_c = soil_c) %>%
  select(-Land_Type) %>%
  mutate(Land_Type = Final_Land_Use) %>%
  right_join(simple_soilC_regions, by = c('GLU_code', 'Land_Type', 'GCAM_region_ID')) %>%
  rename(final_soil_c = soil_c) %>%
  select(-Land_Type, -soilTimeScale.x, -Basin_long_name.x) %>%
  rename(soilTimeScale = soilTimeScale.y, Basin_long_name = Basin_long_name.y) %>%
  na.omit() %>%
  mutate(GCAM_Rate = (final_soil_c - initial_soil_c)/soilTimeScale, Rate_Difference = Exp_Rate - GCAM_Rate, 
         Exp_k = -log(abs(Exp_Rate)*Time +1)/Time,
         GCAM_k = -log(final_soil_c/initial_soil_c)/soilTimeScale, 
         source = 'Post & Kwon'
  ) %>%
  #This next line corrects the sign of Exp_k--we had to take the absolute value to avoid NaNs, so this accounts for that 
  #This did not have to be done for the GCAM_k because the argument inside the natural log will always be positive--it is a positive value divided by another positive value--so sign correction was not necesary. 
  mutate(Exp_k = ifelse(sign(Exp_k) == sign(Exp_Rate), Exp_k*(-1), Exp_k)) -> PostKwon_Comparison
```

Note that, because of how the k value is computed for the experimental data, the sign was often mismatched. The sign of k corresponds to the opposite of the sign of the rate, which had to be corrected manually.  



```{r create_comparison_Wei, echo = FALSE}
#Creating Wei comparison data
Wei %>%
  select(Initial_Land_Use, Final_Land_Use, GLU_code, GCAM_region_ID, OC_decrease, Time) %>%
  na.omit() %>%
  mutate(Land_Type = Initial_Land_Use) %>%
  right_join(simple_soilC_regions, by = c('GLU_code', 'Land_Type', 'GCAM_region_ID')) %>%
  rename(initial_soil_c = soil_c) %>%
  select(-Land_Type) %>%
  mutate(Land_Type = Final_Land_Use) %>%
  right_join(simple_soilC_regions, by = c('GLU_code', 'Land_Type', 'GCAM_region_ID')) %>%
  rename(final_soil_c = soil_c) %>%
  select(-Land_Type, -soilTimeScale.x, -Basin_long_name.x) %>%
  rename(soilTimeScale = soilTimeScale.y, Basin_long_name = Basin_long_name.y) %>%
  na.omit() %>%
  mutate(GCAM_Rate = (final_soil_c - initial_soil_c)/soilTimeScale,
         GCAM_k = -log(final_soil_c/initial_soil_c)/soilTimeScale,
         Exp_k = -log(1/((abs(OC_decrease)/100) +1))/Time,
         source = 'Wei et al'
  ) %>%
  #This next line corrects the sign of Exp_k--we had to take the absolute value to avoid NaNs, so this accounts for that 
  mutate(Exp_k = ifelse(sign(Exp_k) == sign(OC_decrease), Exp_k, Exp_k*(-1))) -> Wei_Comparison
```


\Newline
Once the experimental data was matched to its corresponding GCAM outputs, the two data sets (Post& Kwon and Wei et al.) were merged. Importantly, because the rate data does not exist for Wei et al., the rate data was not included in the aggregated data set. 

```{r Create-Full-Comparison}
Full_Comparison <- bind_rows(
  select(PostKwon_Comparison, -Exp_Rate, -Rate_Difference),
  select(Wei_Comparison, -OC_decrease)
  )
```

Once  the k values were calculated, t-tests and ANOVAs were performed on the data to determine if there was a significant difference of the means of the empirical and GCAM k values. Before the t-tests were performed, the data had to be put in long form and a transition type column was added: 

```{r Merge_Data_and_add_transition}
Full_Comparison %>%
  pivot_longer(cols = Exp_k:GCAM_k,
               names_to = "Type",
               values_to = "k") -> full_long_data 

full_long_data %>% 
  mutate(change = paste(Initial_Land_Use, Final_Land_Use, sep = '')) -> change_long_data
```

We also checked if the variances were equal.

```{r}
vars = var.test(k ~ Type, data = change_long_data)
```
They were not, so we performed some individual Welch's t-tests to check if the aov function was sufficiently robust to handle deviations in variance. If these results matched those from the ANOVAs, we could say that the ANOVA results were not significantly affected by unequal variances.

## Basic characteristics of data, decent number points.

Look at the basic characteristics of the data, and specifically filter
down to basins with at least 3 data points (3 transitions of any kind).
You can make this a little stricter by uncommenting out all the lines with 
`change` - that would say we only consider basins with at least 3 of the same
kind of transition.

```{r}
knitr::kable(head(change_long_data))

change_long_data %>%
  group_by(Basin_long_name, change, Type) %>%
  summarise(n_points  = n()) %>%
  ungroup %>%
  knitr::kable(.)


change_long_data %>%
  group_by(Basin_long_name, 
           #change,
           Type) %>%
  summarise(n_points  = n()) %>%
  ungroup %>%
  filter(n_points >1)  %>%
   mutate(joint_id = paste(Basin_long_name,
                           #change, 
                           sep = '~')) ->
  decent_num_points
knitr::kable(decent_num_points)

change_long_data %>%
  mutate(joint_id = paste(Basin_long_name, 
                          #change,
                          sep = '~')) %>%
  filter(joint_id %in% unique(decent_num_points$joint_id)) ->
  data_for_analysis

knitr::kable(head(data_for_analysis))

print( paste('Basins with at least 3 samples of any transition to validate against:', length(unique(data_for_analysis$Basin_long_name))))
```

## t-tests by basin

By default, R runs welch's t-test, which assumes the variances differ between
groups. 

```{r}

basin_pval_holder <- data.frame()


for (basin in unique(data_for_analysis$Basin_long_name)){
  # print(basin)
  
  data_for_analysis %>% 
    filter(Basin_long_name == basin,
           Type == 'GCAM_k') ->
    gcam_basin
  
  data_for_analysis %>% 
    filter(Basin_long_name == basin,
           Type == 'Exp_k') ->
    exp_basin
  
  pval <- t.test(exp_basin$k, gcam_basin$k, alternative = 'two.sided')$p.value
  
 # print(paste('two-sided alternative, t.test pvalue: ', pval))
 
 basin_pval_holder %>%
   bind_rows(data.frame(basin = basin, p.value = pval), .) ->
   basin_pval_holder
}

```

## interpretation

So regardless of transition type, for basins with at least 3 points of any kind 
of transition recorded, if the p value of the t.test between the experimental
and GCAM distributions of k is greater than 0.05, we fail to reject the hypothesis
that the mean of the distribution of k values from GCAM differs from the mean of 
the distribution of k values from experiment. 

```{r}

print(paste('In:' ,
            length(which(basin_pval_holder$p.value > 0.05)),
            'of', nrow(basin_pval_holder), 
            'basins, we fail to reject the null hypothesis that the means differ.'))
```

So it looks like across 21 of 29 basins with decent amount of data to validate, 
GCAM and experiments, we infer the k-values are comparable, suggesting the 
GCAM approach validates well in these basins

### differing basins

Let's take a look at the 8 basins that differ. 


```{r}
basin_pval_holder %>%
  filter(p.value <= 0.05) %>%
  select(basin) ->
  differing_basins

change_long_data %>%
  filter(basin %in% differing_basins$basin) %>%
  select(initial_soil_c, final_soil_c, Basin_long_name, Type, k, change) %>%
  distinct %>% 
  knitr::kable(.)
```



# anovas

```{r}
summary(aov(k ~ Type * Basin_long_name * change,
               data = change_long_data))
```
Means don't differ if just slice by GCAM vs experiment. They do differ if you 
look across basins (and lumping GCAM vs experiment into the same population in 
each basin) or look across transition types, which I think would make sense? 

They don't differ if you break down by Type and Basin 
(consistent with t-test in each basin having same means for 21 of 29 basins
checked), but it looks like there may be some more significant differences in 
the k-values by transition type any way that is used for grouping distributions.


## Welch's anova

Unfortunately only for one-way anova. SO we can compare the one-way test result
for each predictor via Welch and the equal variances versions. 

```{r}
oneway.test(k ~ Type ,
            data = change_long_data, var.equal = FALSE)

oneway.test(k ~ Type ,
            data = change_long_data, var.equal = TRUE)

oneway.test(k ~ Basin_long_name ,
            data = change_long_data, var.equal = FALSE)

oneway.test(k ~ Basin_long_name ,
            data = change_long_data, var.equal = TRUE)


oneway.test(k ~ change,
            data = change_long_data, var.equal = FALSE)

oneway.test(k ~ change ,
            data = change_long_data, var.equal = TRUE)
```


We do get pretty different results when we do the one-way anova across basins and
transition types with and without the assumption of equal variances. The conclusions
are all the same (there are significant differences, which we expect and also don't
care about because it's not comparing GCAM to observation) but I think that means
we don't want to rely on our three-way anova in the previous section. 

The basin-by-basin t-tests however are fine, and do suggest that the GCAM approach
is broadly consistent with experiments in 21 of 29 basins where data exists in decent quantities.

## maxvar <= 4 * minvar test


```{r}

change_long_data %>%
  group_by(Type) %>%
  summarize(type_var = sd(k) ^2) %>%
  ungroup ->
  type_var

change_long_data %>%
  group_by(Basin_long_name) %>%
  summarize(basin_var = sd(k) ^2) %>%
  ungroup ->
  basin_var

change_long_data %>%
  group_by(Type, Basin_long_name) %>%
  summarize(type_basin_var = sd(k) ^2) %>%
  ungroup ->
  type_basin_var


print(max(type_var$type_var) / min(type_var$type_var))


print(max(basin_var$basin_var) / min(basin_var$basin_var))


print(paste(max(type_basin_var$type_basin_var, na.rm = T), min(type_basin_var$type_basin_var, na.rm = T)))

```

Then, the ANOVAs could be performed. The Akaike information criterion was used to assess which ANOVA best captured the data.

```{r All_Anovas, echo = TRUE, results = 'hide', warning=FALSE}
#Type + Basin 
aov_Full <- aov(k ~ Type + Basin_long_name ,
                      data = full_long_data)
summary(aov_Full)
TukeyHSD(aov_Full)
AIC(aov_Full)


#Type + Change 
aov_Full_change <- aov(k ~ Type + change,
                      data = change_long_data)
summary(aov_Full_change)
TukeyHSD(aov_Full_change)
AIC(aov_Full_change)

#Type + Basin + change
aov_big <- aov(k ~ Type + Basin_long_name + change,
               data = change_long_data)
summary(aov_big)
TukeyHSD(aov_big)
AIC(aov_big)

#Type * Change 
aov_Full_type_by_change <- aov(k ~ Type * change,
                       data = change_long_data)
summary(aov_Full_type_by_change)
TukeyHSD(aov_Full_type_by_change)
AIC(aov_Full_type_by_change)

#Type * Basin 
aov_Full_type_by_basin <- aov(k ~ Type * Basin_long_name,
                               data = change_long_data)
summary(aov_Full_type_by_basin)
TukeyHSD(aov_Full_type_by_basin)
AIC(aov_Full_type_by_basin)

#Type * Basin * Change 
aov_Full_type_by_basin_by_change <- aov(k ~ Type * Basin_long_name * change,
                              data = change_long_data)
summary(aov_Full_type_by_basin_by_change)
AIC(aov_Full_type_by_basin_by_change)
```


# Results
Overall, GCAMs rate and k value outputs matched experimental values. 
```{r k-Histogram-full, echo=FALSE, fig.cap="Histograms showing overlapping GCAM and experimental data for computed k values during land use change. Here, the experimental data is taken from both Post & Kwon and Wei et al.", message=FALSE}
#Plot overlapping k histograms for the different k sources
ggplot() +
  geom_histogram(aes(x = Full_Comparison$Exp_k, fill = Full_Comparison$source), alpha = 0.5, bins = 75) +
  geom_histogram(aes(x = Full_Comparison$GCAM_k,  fill = 'GCAM'), alpha = 0.5, bins = 75) +
  xlab(expression(k~(y^-1))) + ylab('Count') +
  scale_fill_manual(name = "Data Source", 
                    values = c('GCAM'='#e3962b', 'Wei et al' = '#45912c', 'Post & Kwon' = '#3584B0')) +
  theme_light() 
```


Several t-tests were also performed. One was performed on each of the different k value sources (Post & Kwon, Wei et al., and the aggregated data), and another was performed on the Post & Kwon rate data. The difference was significantly significant for the Post & Kwon and Wei et al. data individually, but not for aggregated k values.   
```{r t-test-Post-Kwon}
t.test(PostKwon_Comparison$Exp_k, PostKwon_Comparison$GCAM_k, alternative = 'two.sided')
```

```{r t-test-Wei}
t.test(Wei_Comparison$Exp_k, Wei_Comparison$GCAM_k, alternative = 'two.sided') 

```

```{r t-test-full}
t.test(Full_Comparison$Exp_k, Full_Comparison$GCAM_k, alternative = 'two.sided')
```

Based on the AIC for each ANOVA, the tests accounting for data source by transition type by region was used. These tests showed no significant difference in means based on data source. 

```{r}
summary(aov_Full_type_by_basin_by_change)
```

The TukeyHSD results had to be converted to a dataframe and filtered to be more easily analyzed.

```{r}
type_basin_ty <- TukeyHSD(aov_Full_type_by_basin_by_change, which = 'Type:Basin_long_name')
type_change_ty <- TukeyHSD(aov_Full_type_by_basin_by_change, which = 'Type:change')
basin_change_ty <- TukeyHSD(aov_Full_type_by_basin_by_change, which = 'Basin_long_name:change')

#Convert the Tukey outputs 
as.data.frame(type_basin_ty[1]) %>% 
  arrange(Type.Basin_long_name.p.adj) %>%
  filter(Type.Basin_long_name.p.adj < .01) -> sorted_ty_type_basin

as.data.frame(type_change_ty[1]) %>% 
  arrange(Type.change.p.adj) %>%
  filter(Type.change.p.adj < .01) -> sorted_ty_type_change

as.data.frame(basin_change_ty[1]) %>% 
  arrange(Basin_long_name.change.p.adj) %>%
  filter(Basin_long_name.change.p.adj < .01) -> sorted_ty_basin_change

```


The average of the difference in means between k values was also plotted, grouped by transition type:

```{r Mean_Difference_by_Transition}
#Make the mean graph
Full_Comparison %>%
  mutate(change = paste(Initial_Land_Use, Final_Land_Use, sep = ' to ')) %>%
  mutate(difference = Exp_k - GCAM_k) %>%
  group_by(change) %>%
  summarise(Mean_k = mean(difference), std_dev = sd(difference), Sample_Size = n()) -> mean_difference

ggplot(mean_difference, aes(x = Mean_k, y = change)) +
  geom_point(aes(size = Sample_Size), shape = 15) +
  labs(x = 'k', y = 'Transition Type') +
  geom_errorbar(aes(xmin=Mean_k - std_dev, xmax=Mean_k + std_dev), width=.2,
                position=position_dodge(0.05)) +
  xlim(-0.1, 0.1) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'blue') +
  theme_light() 
```

